`r glue::glue('Initial inferCNV results for {params$library_id}')`

### 2025-04-17 analyses/infercnv-consensus-cell-type/template-notebooks/ewings-infercnv-results.Rmd
Update cell type labeling scheme

@@

@@

# Read SCE and save UMAP coordinates and cell types
umap_df <- readr::read_rds(sce_file) |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  dplyr::select(
    barcodes,
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2
  ) |>
  # TODO: We currently are bringing in cell types from separate files, but in the next
  # data release they will be in the SCE objects
  dplyr::inner_join(
    readr::read_tsv(ewings_tsv) |>
      dplyr::select(barcodes, consensus_annotation, ewing_annotation)
  )
```

@@

Ideally, we'll see that Unknown cells from `cell-type-consensus` and cells labeled as tumor in `cell-type-ewings` will have relatively higher CNV scores compared to other cell types.
In addition, comparing distributions between `all-immune` and `subset-immune` may be helpful to identify the optimal normal reference strategy.

@@

consensus_violin <- ggplot(query_has_cnv_df) +
  aes(
    x = consensus_annotation,
    y = total_cnv_per_cell,
    fill = normal_reference
  ) +
  geom_violin(scale = "width") +
  scale_fill_brewer(palette = "Pastel2") +
  labs(title = "Annotations from cell-type-consensus") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ewings_violin <- ggplot(query_has_cnv_df) +
  aes(
    x = ewing_annotation,
    y = total_cnv_per_cell,
    fill = normal_reference
  ) +
  geom_violin(scale = "width") +
  scale_fill_brewer(palette = "Pastel2") +
  labs(title = "Annotations from cell-type-ewings") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

wrap_plots(consensus_violin, ewings_violin, ncol = 1, guides = "collect") +
  plot_annotation(title = "CNV scores across cell types") &
  theme(legend.position = "bottom")
```

@@

First, as a guide to interpret the UMAP, let's color by the top 10 `cell-type-ewings` cell types.

```{r, fig.width = 6}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = forcats::fct_lump(ewing_annotation, 8)) +
  geom_point(size = 0.1, alpha = 0.3) +
  labs(color = "ewings annotation") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```

@@

Let's also look at the distributions of cell types in clusters.
Is there a trend for which cell types tend to show up in which cluster?
We'll plot this for both the consensus and ewing annotations, considering only the top 7 cell types for visual ease.

```{r fig.width=12}
ggplot(query_has_cnv_df) +
  aes(x = subcluster, fill = forcats::fct_lump(consensus_annotation, 7)) +
  geom_bar() +
  scale_fill_brewer(palette = "Dark2") +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  labs(
    title = "Consensus cell types",
    fill = "annotation"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r fig.width=12}
ggplot(query_has_cnv_df) +
  aes(x = subcluster, fill = forcats::fct_lump(ewing_annotation, 7)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  labs(
    title = "Ewing cell types",
    fill = "annotation"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

>>>>

# immune cell URL to check for ambiguous cell types
immune_url <- "https://raw.githubusercontent.com/AlexsLemonade/OpenScPCA-analysis/refs/heads/main/analyses/cell-type-consensus/references/consensus-immune-cell-types.tsv"

@@

immune_labels <- readr::read_tsv(immune_url) |>
  dplyr::pull(consensus_annotation)

@@

# Read SCE and save UMAP coordinates and cell types
umap_df <- readr::read_rds(sce_file) |>
  scuttle::makePerCellDF(use.dimred = "UMAP") |>
  dplyr::select(
    barcodes,
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2
  )

# TODO: We currently are bringing in cell types from separate files, but in the next
# data release they will be in the SCE objects this could be part of the colData wrangling above.
celltype_df <- readr::read_tsv(ewings_tsv) |>
  # Label cells as:
  # - tumor-immune: Cells labeled as tumor by cell-type-ewings, but immune by cell-type-consensus
  # - tumor: Cells labeled as tumor by cell-type-ewings
  # - their existing ewing_annotation if none of the above
  dplyr::mutate(
    celltype = dplyr::case_when(
      ## tumor-immune
      consensus_annotation %in% immune_labels &
        stringr::str_starts(ewing_annotation, "tumor") ~ "tumor-immune",
      ## tumor
      stringr::str_starts(ewing_annotation, "tumor") ~ "tumor",
      ## default
      .default = ewing_annotation
    )
  ) |>
  dplyr::select(barcodes, celltype)

umap_df <- umap_df |>
  dplyr::left_join(celltype_df)
```

@@

Ideally, we'll see that tumor and Unknown cells from `cell-type-consensus` and cells labeled as tumor in `cell-type-ewings` will have relatively higher CNV scores compared to other cell types.

In addition, it will be useful to:

* See values for cells labeled `tumor-immune`, which were labeled as `tumor` by `cell-type-ewings` but as an immune cell type by `cell-type-consensus`
* Compare distributions between `all-immune` and `subset-immune` which may support identifying the optimal normal reference strategy.

@@

ggplot(query_has_cnv_df) +
  aes(
    x = celltype,
    y = total_cnv_per_cell,
    fill = normal_reference
  ) +
  geom_violin(scale = "width") +
  scale_fill_brewer(palette = "Pastel2") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

First, as a guide to interpret the UMAP, let's color by cell types.

```{r, fig.width = 6}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = celltype) +
  geom_point(size = 0.1, alpha = 0.3) +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```

@@

Let's also look at the distributions of cell types in clusters.
Is there a trend for which cell types tend to show up in which cluster?

```{r fig.width=12}
ggplot(query_has_cnv_df) +
  aes(x = subcluster, fill = celltype) +
  geom_bar() +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(
    vars(normal_reference),
    scales = "free_x"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
